---
title: "The Academy Awards ... Oscars"
subtitle: "Discrete choice models"
author: "Bikramjit Das"
date: "2025"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: yes
    toc_depth: 2
    number_sections: yes
---
# Introduction

The Academy (of Motion Picture Arts and Sciences) Awards of the Oscars is perhaps the most well-known honor for  English language commercial cinema. Winning an Oscar leads to many better deals for the parties involved (actors, directors, producers, ...). Is it possible to predict the Oscars awards with any degree of accuracy? We follow the methodology of a multinomial choice model. 

Fur further details, analysis and recent updates, you can check the following webpage:  
  
  <https://iainpardoe.com/oscars/> 


## Read the data
```{r Read-data, message=FALSE, warning=FALSE, paged.print=FALSE, results='hide'}
oscars<- read.csv("oscars.csv")
str(oscars)
#View(oscars)
summary(oscars)
```

## Data attributes
The following attributes are observed in the dataset:

 * `Year`: Movie year  
 
 * `Name`: Nominee name  
 
 * `PP`: Indicator for picture  
 
 * `DD`: Indicator for director  
 
 * `MM`: Indicator for lead actor (male)  
 
 * `FF`: Indicator for lead actress (female)  
 
 * `Mode`: Alternative (choice) number (1 to 5 here)  
 
 * `Ch`: 1 = Winner, 2= No  
 
 * `Movie`: Movie name  
 
 * `Nom`: Number of Oscar nominations  
 
 * `Pic`: Picture nomination  
 
 * `Dir`: Director nomination  
 
 * `Aml`: Lead actor (male) nomination  
 
 * `Afl`: Lead actress (female) nomination  
 
 * `PrN`: Total previous acting/directing nominations  
 
 * `PrW`: Total previous acting/directing wins  
 
 * `PrNl`: Previous lead acting nomination  
 
 * `PrWl`: Previous lead acting wins  
 
 * `Gdr`: Golden Globe drama winner  
 
 * `Gmc`: Golden Globe musical or comedy winner  
 
 * `Gd`: Golden Globe director winner  
 
 * `Gm1`: Golden Globe drama actor winner  
 
 * `Gm2`: Golden Globe musical or comedy actor winner  
 
 * `Gf1`: Golden Globe drama actress winner  
 
 * `Gf2`: Golden Globe musical or comedy actress winner  
 
 * `PGA`: Producers guild winner  
 
 * `DGA`: Directors guild winner  
 
 * `SAM`: Screen actors guild actor winner  
 
 * `SAF`: Screen actors guild actress winner  
 
 * `Age`: Actor/actress age in movie year  
 
 * `Length`: Run time  
 
 * `Days`: Days between release date and Oscars ceremony  
 

To help in multinomial logit modeling we convert `Ch`: 0 = No, 1 = winner  
 
```{r}
oscars$Ch <- 2-oscars$Ch
```



# Using pre-Oscar data

The data set consists of nominees and winners in four categories- *Best Picture, Best Director, Best Actor* and *Best Actress*.  To predict the winners in a given year and category, we can make use of data available prior to the **Oscar** awards.

For example, information on the number of nominations that a movie gets in the **Oscars**, if the movie, actors, director won awards earlier in the season such as **Golden Globes**, have the actors, directors been nominated earlier (body of work).

Let us look at a few criteria that can be used to predict ht Oscars.

## Question 1: Best Picture: does nomination in other categories matter?

Does the winner of the *Best Picture* have more nomination in Oscar categories as compared to the losing nominees?

```{r}
tapply(oscars$Nom[oscars$PP==1],oscars$Ch[oscars$PP==1],mean)
```
In the data set, the winning movies on average have 9.526 nominations compared to the 6.78 for losing nominees.

```{r}
tapply(oscars$Nom[oscars$PP==1],oscars$Ch[oscars$PP==1],var)
```
Variance is comparable across observations.


```{r}
t.test(oscars$Nom[oscars$PP==1 & oscars$Ch==1],oscars$Nom[oscars$PP==1 &oscars$Ch==0], alternative = c("greater"))
```

The `p-value` is very low. Hence we can reject the null hypothesis that the winning picture has equal or fewer nominations than losing nominees.

## Question 2: Best Picture: does Best Director nomination matter?

Do the Best Picture winners also receive nominations for Best Directors?

```{r}
table(oscars$Dir[oscars$PP==1 & oscars$Ch==1])
```

Of the 57 best picture winners, only 1 of them did not get a best director nomination.
```{r}
which(oscars$Dir==0 & oscars$PP==1 & oscars$Ch==1)
```
Row 362.

To find the name of the movie and the year
```{r}
oscars[which(oscars$Dir==0 & oscars$PP==1 & oscars$Ch==1), c("Year","Name")]
```
This movie is *Driving Miss Daisy* which did not get best director nomination but won best picture.


## Question 3: Best Actor/Actress: does Best Picture nomination matter?

Do the Best Actor and Best Actress winners have nominations for movies in the Best Picture category?
```{r}
table(oscars$Pic[oscars$MM==1 & oscars$Ch==1])
```
Out of 57 movies where the actor won the best actor award, 14 were not nominated in best picture category.

```{r}
table(oscars$Pic[oscars$FF==1 & oscars$Ch==1])
```
Out of 58 movies where the actress won the best actress award, 23 were not nominated for best picture.

Surprisingly there is one extra winner in the Best Actress category.

```{r}
oscars$Year[oscars$FF==1 & oscars$Ch==1]
```

We can see that in 1968 there are two awards.

```{r Shared_actress_award, results='hide'}
subset(oscars, Year==1968 & FF==1)
```
Katherine Hepburn for Lion in Winter and Barbara Streisand for Funny Girl shared the Best Actress award with 3030 votes each.


## Question 4: Does winning other awards matter (all categories)?

The Golden Globe awards are awarded typically one to two months before the Oscar awards. The award is bestowed by 93 members of the Hollywood Foreign Press Association. The award has been given every year since 1944.

The Directors Guild of America has been awarding Best Motion Picture Director since 1949, Producer Guild of America has been awarding Best Producing effort since 1989. Since 1994, Screen Guild has been awarding Best Male Actor and Female Actor in a leading role. These awards are also typically given before the Oscars and can be used as an indicator of chance of success.

Since 1951, this award has been given before the Oscars hence yielding some possible predictive power in the model.

In the data set, the DGA award is used till 1989 and then PGA for coding the Best Picture award.


### Best Picture: Golden Globe  
Do the Golden Globe awards help predict the Oscars (Best Picture)? 

```{r}
table(oscars$Gdr[oscars$PP==1 & oscars$Ch==1] + oscars$Gmc[oscars$PP==1 & oscars$Ch==1])
```
Out of the 57 Best Picture Awards given between 1951 and 2006, 39 won the Best Golden Globe picture award, i.e., 39/57 = 0.684 proportion also Golden Globe winners.

### Best Director: Golden Globe 
What about Best Director?

```{r}
table(oscars$Gd[oscars$DD==1 & oscars$Ch==1])
```
Best Director: 31/57 = 0.543 proportion are Golden Globe winners.

### Best Actor: Golden Globe 
What about Best Actor?
```{r}
table(oscars$Gm1[oscars$MM==1 & oscars$Ch==1]+oscars$Gm2[oscars$MM==1 & oscars$Ch==1]) 
table(oscars$SAM[oscars$MM==1 & oscars$Ch==1]) 
```

Best Actor: 42/57 = 0.736 (Golden Globe)
Best Actor: 10/57 = 0.175 (Screen Actors Guild)

### Best Actress: Golden Globe 
What about Best Actress?
```{r}
table(oscars$Gf1[oscars$FF==1 & oscars$Ch==1] + oscars$Gf2[oscars$FF==1 & oscars$Ch==1])
table(oscars$SAF[oscars$FF==1 & oscars$Ch==1]) 
```

Best Actress: 40/58 = 0.689 (Golden Globe)
Best Actress: 10/58 = 0.172 (Screen Actors Guild)


## Best Actor/Actress: Previous nomination

What is the effect of having nominations in the previous years on winning in the current year?
What is the effect of having won awards in the previous years for Oscars to winning in a current year?


**Best Actor**

```{r}
table(oscars$PrNl[oscars$MM==1], oscars$Ch[oscars$MM==1])
table(oscars$PrWl[oscars$MM==1], oscars$Ch[oscars$MM==1])
```

### Effect of previous nomination 

27/(111+27) = 0.195  

30/147= 0.207  

 * About 19.5% of Best Actor nominees with no previous lead nomination won.  
 * About 20.7% of Best Actor nominees with one or more previous lead nominations won.

### Effect of previous win

51/(176+51) = 0.224  

6/(6+41+11) = 0.103  



* 22% of Best Actor Oscar nominees with no previous lead actor wins won the Oscars while it is 10% for actors with a previous win.  


**Best Actress**
```{r}
a<-table(oscars$PrNl[oscars$FF==1], oscars$Ch[oscars$FF==1])
b<-table(oscars$PrWl[oscars$FF==1], oscars$Ch[oscars$FF==1])
a
b
```

### Effect of previous nomination 

32/(110+32) = 0.225  

25/142= 0.176   


 * About 22.5% of Best Actress nominees with no previous lead nomination won.  
 
 * About 17.6% of Best Actress nominees with one or more previous nominations won.
 
 
### Effect of previous win

47/(164+47) = 0.223  

11/(56+7+9+1+1) = 0.149  

 * 22% of Best Actress Oscar nominees with no previous lead actress wins won the Oscars while it is 15% for actresses with a previous win.

# Discrete choice modeling

Now we use a Discrete Choice Model to predict Oscar winners.

## Load `mlogit`

Load the package from Multinomial Logit
```{r install mlogit}
#options(repos="https://cran.rstudio.com" )
#install.packages("mlogit")
```

```{r results='hide'}
library(mlogit)
```


## Create dataframes

Create dataframes for *Best Picture, Best Director, Best Male Actor, Best Female Actor*
```{r}
oscarsPP <- subset(oscars, PP==1)
oscarsDD <- subset(oscars, DD==1)
oscarsMM <- subset(oscars, MM==1)
oscarsFF <- subset(oscars, FF==1)
```

```{r results='hide'}
str(oscarsPP)
#View(oscarsPP)
```

*Best Picture:* 285 observations

Best Picture winner given by `Ch=1` (winner), `Ch=0` for losing nominee  

__Possible Predictors in the data set__

 * `Nom`: no. of Oscar nominations  
  
 * `Dir`: 1 = director nominated for Oscar that year, 0 otherwise

 * `GG`: `Gmc + Gdr` = 1 if movie wins golden globe, 0 otherwise

 * `Aml`: Lead actor nomination

 * `Afl`: Lead actress nomination

 * `PGA`: Producers Guild Award

 * `Days`: Days between release and Oscars ceremony

 * `Length`: Run time of movie

We use this to define a new variable that captures if a movie won a Golden Globe for best picture.

```{r}
oscarsPP$GG <- oscarsPP$Gmc + oscarsPP$Gdr
```

### Read data set to conduct multimomial logit

Now we use the data frame from 1944 to 2006 to develop the logit model.

```{r}
D1 <- mlogit.data(subset(oscarsPP, Year <=2006), choice="Ch", shape="long", alt.var = "Mode")
```

This creates a data set for applying the `mlogit` function where `choice` is a variable indicating the choice mode (here `Ch`), `shape` is the shape of the data frame (here "long" since each row is an alternative) and `alt.var` is the name of the variable containing the alternative index (here `Mode`). We use `shape="wide"` when there is one row for all the alternatives.

## Model 1

```{r}
MPP1 <- mlogit(Ch~Nom+Dir+GG+Aml+Afl+PGA+Days+Length-1, data=D1)
```

This fits a conditional logit model where `Ch` is the response. The -1 is used to address the fact that in this fit, we do not want the intercept to be estimated. Note that across the five alternatives in different years, it is not comparable and hence we should not introduce alternate specific estimates here.

```{r}
summary(MPP1)
```

`Nom`, `Dir`, `GG` and `PGA` are the most significant variables in the fit. 

The length of movies, the number of days it was released before the Oscars, whether a lead actor got nominated for the best picture are less significant. Note that the variables `Aml` and `Afl` are included in the `Nom` variable (multi-collinearity). 

## Model 2

Consider a simple model using only the variables:


* `Nom` (No, of Oscar nominations), `Dir` (Director nomination), `GG` (Golden Globe winner), `PGA` (Producer Guild winner).  

* Output: `Ch`  

```{r}
MPP2 <- mlogit(Ch~Nom+Dir+GG+PGA-1, data=D1)
summary(MPP2)
```

The model obtained is as follows:

$$\mathbf{P}(\text{Movie $k$ wins the best picture}) = \frac{exp\{0.21Nom_k + 2.63Dir_k + 0.69GG_k + 1.84PGA_k\}}{\sum\limits_{l=1}^K exp\{0.21Nom_l + 2.63Dir_l + 0.69GG_l + 1.84PGA_l\}}$$

## Model 2: Quality of Fit
```{r}
LL0 <- 56*log(1/5)
LLbeta1 = as.numeric(MPP1$logLik)
LLR1 = 1-LLbeta1/LL0
LLR1
p1=8
AIC1 = -2*LLbeta1+2*p1
AIC1

LLbeta2 = as.numeric(MPP2$logLik)
LLR2 = 1-LLbeta2/LL0
LLR2
p2=4
AIC2 = -2*LLbeta2+2*p2
AIC2

```

Likelihood  Ratio index for the second model is $\rho = 1 - \frac{LL(\hat{\beta})}{LL(0)}=$ `r LLR2`. 


The base log-likelihood is $LL(0) = 56 \log (1/5)=$ `r 56*log(0.2)` where each alternative is picked equally likely assuming $\beta_1=\beta_2=\ldots=\beta_p=0$ and no. of choice tasks = 56.

Moroever $LL(\beta)= $ `r as.numeric(MPP2$logLik)`.

Then $AIC =  -2 LL(\hat{\beta})+2 p=$ `r AIC2`.  

Note that if we use the expanded model with variables `Nom`, `Dir`, `GG`, `Aml`, `Afl`, `PGA`, `Days`, `Length`, the `AIC` value = `r AIC1`` (larger). 

# Prediction

Now we predict the outcome for year 2007 . First we need to structure the data and then predict (using Model 2).
```{r}
D1_new <- mlogit.data(subset(oscarsPP, Year==2007), choice="Ch", shape="long", alt.var="Mode")
Predict2 <- predict(MPP2, newdata= D1_new)
Predict2
```


```{r results='hide'}
oscarsPP[oscarsPP$Year==2007 & oscarsPP$Mode==which.max(Predict2),]
subset(oscarsPP, Year==2007)
```

 * Winner for Oscars 2007 best picture: No country for Old Men.
 
This movie had the highest predicted probability from the model.
8 Nominations, director nominated for best picture, won PGA but not Golden Globe award.

## Surprise Winners

```{r results='hide'}
D <- mlogit.data(oscarsPP, choice = "Ch", shape="long", alt.var="Mode")
M <- mlogit(Ch~Nom+Dir+GG+PGA-1, data=D)
P <- predict(M, newdata=D)
Pred <- as.vector(t(P))
oscarsPP$Pred <- Pred
cbind(oscarsPP$Pred[oscarsPP$Ch==1],oscarsPP$Name[oscarsPP$Ch==1])
#oscarsPP %>% select(Year,Movie,Ch,Pred)
subset(oscarsPP, oscarsPP$Year == 2004)
```

For example, in the year 2004, Million Dollar Baby won the Best Picture with predicted probability of 0.02, though based on the model, The Aviator was the overwhelming favorite with predicted probability of 0.90.


## A list of predictions and real winners

We make a list of predicted versus winners below. Note that we use the the data set until 2006 for creating the prediction model.
```{r}
PPwin=apply(P,1,which.max)
PPyw=cbind(2007:1951,PPwin)
lists=character(0)
for (i in 1:57)
  lists=rbind(lists,c(PPyw[i,1],as.character(oscarsPP[oscarsPP$Year==PPyw[i,1] & oscarsPP$Mode==PPyw[i,2],c("Name")]),round(P[i,PPyw[i,2]],digits=2),as.character(oscarsPP[oscarsPP$Year==PPyw[i,1] & oscarsPP$Ch==1,c("Name")])))

colnames(lists)=c("Year","Predicted","Probability","Winner")
noquote(lists)
```



## Best Actors

Next we look at the male actors.

Here we predict the next year using previous year's data. Prediction starts from 1960.

```{r}
Fail <- 0
Predict <- NULL
coefficients <- NULL  # reserved keyword for null object in R (undefined)
for(i in 1960:2006){
  D <- mlogit.data(subset(oscarsMM, Year<=i), Choice="Ch", shape="long", "alt.var"="Mode")
  M <- mlogit(Ch~Pic+Gm1+Gm2+PrNl+PrWl-1, data=D)
  coefficients <- rbind(coefficients, M$coefficients)
  D1 <- mlogit.data(subset(oscarsMM, Year == (i+1)), choice="Ch", shape="long", alt.var="Mode")
  P1 <- predict(M, newdata=D1)
  Predict <- rbind(Predict, P1)
  Fail <- Fail + as.logical(which.max(P1) - which.max(subset(oscarsMM, Year== (i+1) )$Ch))
}
Predict
Fail
```

```{r}
D_all <- mlogit.data(subset(oscarsMM), Choice="Ch", shape="long", "alt.var"="Mode")
M_all <- mlogit(Ch~Pic+Gm1+Gm2+PrNl+PrWl-1, data=D_all)
summary(M_all)
```

Total number of `fails = 14` out of 57 where Fail corresponds to best actor being someone who the model did not predict with the highest probability.   

Note you can also check from the full model result that `PrNl` does not seem to be significant in predicting winners as discussed earlier.

# Conclusion

Predicting Oscar winners is important in many ways:

 * Many news magazines and media have their own predictions from movie experts in the area  
 * Using quantitative models provides an alternate approach to predict this winner.  
 
This remains an active field for analytics techniques in the movie industry.